---
title: 'Using Machine Learning Algorithms to Predict the likelihood of PCOS based on Demographic, Clinical and Lifestyle Factors'
author: "Leticia Salazar"
date: "May 16, 2023"
output:
  pdf_document:
    latex_engine: xelatex
    toc: yes
  html_document:
    toc: yes
    toc_float: yes
    number_sections: no
    theme: united
    highlight: kate
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

$~$

## Abstract:

<<<<<<< HEAD
Polycystic Ovary Syndrome (PCOS) stands as a prevalent endocrine disorder affecting women of reproductive age worldwide, presenting a confluence of hormonal imbalances, reproductive irregularities, and potential metabolic complications. Characterized by irregular menstrual cycles, hyperandrogenism, and polycystic ovaries, PCOS poses multifaceted challenges that extend beyond reproductive health, encompassing metabolic disturbances and psychological implications. This complex syndrome, rooted in intricate interplays of genetic predispositions, hormonal dysregulation, and environmental factors, manifests variably among affected individuals. The clinical landscape of PCOS often requires a comprehensive, multidisciplinary approach, incorporating lifestyle modifications, pharmacological interventions, and personalized treatments to address symptoms and reduce associated health risks. Despite ongoing research efforts, elucidating the precise etiology and optimal management strategies for PCOS remains a dynamic area of exploration in contemporary medicine. This research project aims investigates the current studies on PCOS diagnosis, assessing the effectiveness of various machine learning algorithms including Linear Regression (LM), Logistic Regression, Decision Tree (rpart), Random Forest (randomForest), Gradient Boosting Machines (xgboost or gbm), Support Vector Machines (e1071), Neural Networks (neuralnet), and K-Nearest Neighbors (kknn or class). From my research 

=======
Polycystic Ovary Syndrome (PCOS) stands as a prevalent endocrine disorder affecting women of reproductive age worldwide, presenting a confluence of hormonal imbalances, reproductive irregularities, and potential metabolic complications. Characterized by irregular menstrual cycles, hyperandrogenism, and polycystic ovaries, PCOS poses multifaceted challenges that extend beyond reproductive health, encompassing metabolic disturbances and psychological implications. This complex syndrome, rooted in intricate interplays of genetic predispositions, hormonal dysregulation, and environmental factors, manifests variably among affected individuals. The clinical landscape of PCOS often requires a comprehensive, multidisciplinary approach, incorporating lifestyle modifications, pharmacological interventions, and personalized treatments to address symptoms and reduce associated health risks. Despite ongoing research efforts, elucidating the precise etiology and optimal management strategies for PCOS remains a dynamic area of exploration in contemporary medicine. This research project investigates the current studies on PCOS diagnosis, assessing the effectiveness of various machine learning algorithms including Linear Regression (LM), Logistic Regression, Decision Tree (rpart), Random Forest (randomForest), Gradient Boosting Machines (xgboost or gbm), Support Vector Machines (e1071), Neural Networks (neuralnet), and K-Nearest Neighbors (kknn or class).
>>>>>>> 5817706a4d5d7c4fb8bdab52d62f5a19693273c3

$~$

## Key words:

<<<<<<< HEAD
Polysyctic Ovary Syndrome, Polycystic Ovarian Syndrome, PCOS, Women's Health, Machine Learning, Linear Regression, Logistic Regression, Decision Tree, Random Forest, Gradient Boosting, Super Vector Machine, Neural Networks, K-Nearest Neighbors
=======
Polysyctic Ovary Syndrome, Polycystic Ovarian Syndrome, PCOS, Women's Health, Machine Learning, Linear Regression, Logistic Regression, Decision Tree, Random Forest, Gradient Boosting, Super Vector Machine, Neural Networks, K-Nearest Neighbors, KNN
>>>>>>> 5817706a4d5d7c4fb8bdab52d62f5a19693273c3

$~$

## The Problem:

Polycystic Ovary Syndrome (PCOS) is a hormonal imbalance disorder affecting women of reproductive age. Determining the precise global count of women affected by PCOS poses challenges due to many cases remaining undiagnosed. However, the World Health Organization (WHO) estimates that approximately 3.4% of women are affected [1]. While this percentage might seem relatively small, considering that women constitute 49.7% of today’s population [2], nearly 13% of them fall within the reproductive age bracket [3]. This data suggests that approximately 17.5 million women report suffering from PCOS. Beyond the physical and emotional toll PCOS exacts, it also disrupts ovarian function, leading to challenges in maintaining a healthy menstrual cycle and can result in the formation of cysts, ultimately impacting fertility. It's important to note that the prevalence of PCOS varies by region and ethnic groups, with some studies suggesting higher rates of PCOS in certain populations [4]. Early identification of risk factors associated with PCOS can assist in timely interventions and lifestyle adjustments. Tailoring suggestions or treatments based on individualized risk profiles has the potential to enhance patient outcomes. Employing predictive models can play a pivotal role in increasing awareness regarding PCOS risk factors and preventive measures. However, limitations may exist in accessing comprehensive and varied datasets containing accurate demographic, clinical, and lifestyle data. It is crucial to ensure the model's transparency and interpretability to facilitate well-informed decisions and recommendations. Additionally, it's imperative that the model demonstrates proficiency across diverse demographic groups and populations. Addressing these challenges involves the development of a robust predictive model using machine learning techniques. Such an endeavor holds the promise of significantly contributing to the identification of individuals at risk of PCOS, thereby enabling early interventions and guiding personalized healthcare strategies for improved management of the condition. This project will investigate publicly available datasets that will be used to develop machine learning models that predicts the probability or risk of an individual having or developing PCOS based on demographic information (age, ethnicity, geographical location), clinical data (hormonal levels, BMI, menstrual irregularities), and lifestyle factors (dietary habits, exercise routine, stress levels).

$~$

### Objectives:

Construct predictive models using machine learning techniques that utilize a dataset comprising demographic, clinical, and lifestyle variables as features and PCOS diagnosis as the target variable. Machine learning algorithms have shown promise in advancing our understanding of the disease and improving its diagnosis and treatment. 

I anticipate answering the following questions with my data:

1.	Are there commonalities women with and without PCOS have that can be easily dismissed as normal?

2.	Are there differences for women of different race/ethnic background when it comes to having PCOS? What about women without PCOS?

3.	What is the likelihood of a woman developing PCOS based on her age, ethnicity, and BMI history?

4.	Can we predict the risk of insulin resistance, diabetes, and cardiovascular disease in women with PCOS based on their medical history, hormone levels, and lifestyle factors?

5.	Can we predict the likelihood of successful pregnancy outcomes in women with PCOS based on their age, weight, hormone levels, and treatment history?

6.	Can we predict the long-term health outcomes and quality of life of women with PCOS based on their age, lifestyle factors, hormone levels, and treatment history?

$~$

## Literature Review:

<<<<<<< HEAD
For this project, my emphasis was on discovering literature reviews that validate the relevance of the dataset utilized alongside the chosen machine learning algorithms. In pursuit of this goal, I've come across various articles that make reference to the Kaggle dataset. These articles investigate studies associated with PCOS, presenting findings aimed at providing a deeper comprehension of the dataset I'm analyzing.
=======
For this project, my emphasis was on discovering literature reviews that validate the relevance of the dataset utilized alongside the chosen machine learning algorithms. In pursuit of this goal, I identified several articles outlined below, including studies related to PCOS and its findings in order to better understand the data I am working with.

1. **A Critical Study of Polycustic Ovarian Syndrome (PCOS) Classification Techniques**: The article discusses the application of data mining techniques in diagnosing Polycystic Ovarian Syndrome (PCOS) in women. It emphasizes the use of classification methods such as Naïve Bayes, Decision Trees, and Neural Networks to predict the likelihood of PCOS based on gathered data. The study uses a dataset obtained from a survey conducted among 119 women aged 18 to 22, focusing on lifestyle and dietary habits. Different classification algorithms are applied and compared for accuracy, with Naïve Bayes demonstrating the highest accuracy of 97.65%. The conclusion suggests integrating clinical reports and ultrasound images to further enhance PCOS prediction and assist in early medical intervention. The article acknowledges contributions from medical professionals and the university in facilitating data collection and guidance throughout the project.

2. **A Review: Brief Insight into Polycustic Ovarian Syndrome**: The article discusses Polycystic Ovarian Syndrome (PCOS), an endocrine disorder impacting women due to ovarian cysts, anovulation, and hormonal imbalances. Around 116 million women worldwide are affected by PCOS, influenced by factors like genetics, lifestyle, and obesity. The disorder's pathophysiology involves hormonal dysfunction, insulin resistance, and excess androgen leading to complications like diabetes and endometrial cancer. PCOS is linked to various risk factors such as genetics, neuroendocrine issues, lifestyle choices, and obesity. The condition leads to hormonal imbalances, particularly increased androgens, which affect folliculogenesis and ovarian function. Insulin resistance, often observed in PCOS, further exacerbates hormonal disruptions and contributes to metabolic issues like diabetes and cardiovascular problems. Treatment options for PCOS mainly focus on managing symptoms. Medications like oral contraceptives, antiandrogens, insulin sensitizers (e.g., metformin, thiazolidinediones), and ovulation-inducing agents (e.g., clomiphene citrate, letrozole) are used to address specific symptoms such as irregular menstruation, hirsutism, and infertility. Lifestyle interventions including diet and exercise play a crucial role in managing PCOS-related symptoms. The article concludes that while various treatments target PCOS symptoms, there's no singular cure. Lifestyle modifications, in addition to pharmacological interventions, are vital for managing the condition's impact on a patient's health. Further research into the pathophysiology and potential herbal or alternative treatments is recommended for improving long-term outcomes in PCOS patients.

3. **A Novel Approach for Polycystic Ovary Syndrome Prediction Using Machine Learning in Bioinformatics**: The article focuses on Polycystic Ovary Syndrome (PCOS), a disorder affecting women's reproductive health due to hormone imbalances, leading to complications such as infertility and pregnancy issues. It discusses the necessity of early detection and proposes a novel approach using advanced machine learning techniques for PCOS prediction. The study employs a dataset with clinical and physical parameters, introducing a novel feature selection technique called CS-PCOS based on optimized chi-squared mechanism. Ten machine learning models are compared, with Gaussian Naive Bayes (GNB) showing superior performance, achieving 100% accuracy, precision, recall, and F1-scores with minimal computational time. The research identifies key factors like prolactin levels, blood pressure, thyroid stimulating hormone, relative risk, and pregnancy as significant in PCOS prediction. It also details exploratory data analysis methods and concludes by suggesting future enhancements such as expanding the dataset and applying deep learning techniques for PCOS prediction.

4. **A Study of Hormonal and Anthropometric Parameters in Polycystic Ovarian Syndrome**: 

5. **Clinical Characteristics of Polycystic Ovary Syndome in Indian Women**: The article investigates the clinical features of Polycystic Ovary Syndrome (PCOS) in Indian women, focusing on identifying different phenotypes of the syndrome. It examines the prevalence of markers like acanthosis nigricans (AN) linked with insulin resistance, obesity, hirsutism, and hypothyroidism in PCOS women. Conducted over 18 months, the study included 120 newly diagnosed PCOS patients and highlighted the following:

  * PCOS Occurrence: The research found PCOS in both obese and non-obese women, with AN and hirsutism occurring in an equal proportion of patients. Hormonal dysfunctions in PCOS manifested together or independently.

  * Infertility: Infertility due to anovulation is a complication of PCOS. However, infertility risks did not significantly differ between obese and non-obese PCOS women.

  * Obesity: While obesity is common in PCOS, the study noted that PCOS occurred in women with normal BMI as well. It found that obesity was not essential to diagnose PCOS.

  * Acanthosis Nigricans: AN, a marker of insulin resistance, was present in a significant proportion of PCOS women, especially among the obese, indicating a higher risk of insulin resistance at comparatively lower BMI levels among Indian PCOS women.

  * Hirsutism and Other Hyperandrogenic Features: The study observed hirsutism, acne, and baldness among PCOS women. While more obese women had hirsutism, there was no direct correlation between hirsutism and obesity. However, both AN and hirsutism were present in a substantial number of patients, especially among the obese, suggesting a potential link between insulin resistance and hyperandrogenism.

The research aimed to identify clinical features that can help subgroup PCOS women based on endocrinological malfunctions for more appropriate treatment selection. It emphasized the need for tailored treatment modalities based on the specific subgroup of PCOS.

6. **Comparative Analysis of Artificial Intelligence in the Diagnosis of Polycystic Ovary Syndrome**: The article explores Polycystic Ovary Syndrome (PCOS) and the challenges in its diagnosis due to its variable symptoms. It highlights the use of Artificial Intelligence (AI), specifically machine learning models, to improve PCOS diagnosis accuracy. Various machine learning techniques like Convolutional Neural Networks (CNN), Artificial Neural Networks (ANN), Support Vector Machines (SVM), Decision Tree (DT), and K-Nearest Neighbors (KNN) were compared for diagnosing PCOS using datasets like ovarian ultrasound images, tongue images, and pulse parameters. The findings revealed that CNN models performed best, achieving accuracies ranging from 85% to 98.12% in different studies. They effectively classified ovarian ultrasound images and detected cysts for PCOS diagnosis. SVMs also showed high accuracies, particularly in tongue image analysis and ultrasound image detection, achieving around 94% accuracy. KNN and Decision Tree models achieved accuracies of 91% and 81%, respectively. The conclusion emphasizes that AI techniques, particularly CNN models, hold promise in improving PCOS diagnosis. However, further advancements are needed through larger datasets, feature selection improvements, and model performance evaluation to enhance the accuracy of PCOS diagnosis. Utilizing ensemble machine learning approaches could also aid in achieving better classification accuracy for PCOS diagnosis.

7. **Comparative Analysis of Machine Learning Algorithms in Diagnosis of Polycystic Ovarian Syndrome**:

8. **DHEA, DHEAS and PCOS**: The article delves into the complex relationship between excess adrenal precursor androgen (APA) production and Polycystic Ovary Syndrome (PCOS) in women. It notes that around 20-30% of women with PCOS demonstrate increased levels of APA, specifically using DHEAS (dehydroepiandrosterone sulfate) as an indicator. However, the direct cause of this excess in PCOS remains unclear. The study discusses observations in patients with inherited APA excess, suggesting that it can lead to a PCOS-like condition. It explores the role of adrenal steroidogenesis exaggeration in women with PCOS, indicating that while it's not a result of overt hypothalamic-pituitary-adrenal axis dysfunction, it might reflect a generalized and inherited exaggeration in androgen biosynthesis. It highlights that while various factors like obesity, insulin, glucose levels, and ovarian secretions might play a role, they have limited influence on increased APA production observed in PCOS. Moreover, genetic studies have shown substantial heritabilities of APAs, particularly DHEAS, but the identified genetic variations only account for a small portion of this inheritance. Several unanswered questions remain, prompting further research. These include understanding the implications of APA excess in PCOS, especially in terms of metabolic dysfunction and cardiovascular risk, exploring the role of emerging endocrine signals, and conducting genetic studies with larger sample sizes to unravel the complex genetic and epigenetic underpinnings of APA production in women with and without PCOS.

This article examines the effect of excess adrenal precursor androgen (APA) production on women with PCOS. The extra-adrenal factors, including obesity, insulin and glucose levels, and ovarian secretions, play a limited role in the increased APA production observed in PCOS.

9. **Early identification of PCOS with commonly known diseases: Obesity, diabetes, high blood pressure and heart disease using machine learning techniques**:

10. **Exercise in polycystic ovarian syndrome: An Evidence-based review**:

11. **Genetic Basis of Polycystic Ovary Syndrome (PCOS) Current Perspectives**:

12. **Hyperparameter Optimization for Machine Learning Models Based on Bayesian Optimization**:

13. **PCOS Detect using Machine Learning Algorithms**:

14. **Phenotype and metabolic profile of South Asian Women with polycystic ovary syndrome (PCOS): results of a large database from a specialist Endocrine Clinic**:

15. **Polycystic Ovarian Syndrome (PCOS) awareness amoing young women of central India**:

16. *Prediction of Polycystic Ovarian Syndrome with Clinical dataset using a novel hydrid data mining classification technique***:

17. **Racial and ethnic differences in the polycystic ovary syndrome metabolic phenotypes**:

18. **SPOSDS: A Smart Polycystic Ovary Syndrome diagnostic system using machine learning**:









The following articles will be resourceful for this project:
>>>>>>> 5817706a4d5d7c4fb8bdab52d62f5a19693273c3

The literature reviews collectively delve into various aspects of Polycystic Ovarian Syndrome (PCOS), employing diverse methodologies and approaches for diagnosis, classification, understanding clinical manifestations, and proposing potential treatments. Researchers across these studies have primarily utilized data mining, machine learning, and clinical investigations to address the complexity of PCOS. Below are the key characteristics, achievements, advantages, and drawbacks across these reviews:

<<<<<<< HEAD
##### Common Focus Areas:
Multiple studies emphasize the use of machine learning algorithms, such as Naïve Bayes, Decision Trees, Artificial Neural Networks, Support Vector Machines, and ensemble methods, for PCOS diagnosis. They explore the accuracy and predictive power of these models using diverse datasets, including clinical, lifestyle, and physiological parameters. Investigations into clinical parameters and anthropometric measures aim to identify potential predictors or indicators of PCOS, such as hormonal imbalances, insulin resistance, metabolic traits, obesity, and associated risks like infertility and cardiovascular issues. Studies examine the diverse clinical presentations and phenotypic variations within PCOS, shedding light on how different subgroups may manifest the syndrome and respond to various treatments.

##### Achievements:
Machine learning models, particularly those utilizing Convolutional Neural Networks (CNNs) and ensemble methods, have shown high accuracy in diagnosing PCOS. These models effectively utilize diverse datasets, ranging from ultrasound images to clinical parameters. Research has identified several potential predictive factors for PCOS, including hormonal markers, lifestyle attributes, and metabolic indicators, offering insights into early detection and tailored treatment strategies. Studies evaluating PCOS awareness among women have highlighted the importance of education and awareness campaigns in enhancing understanding and facilitating early diagnosis.
=======


>>>>>>> 5817706a4d5d7c4fb8bdab52d62f5a19693273c3

##### Advantages:
Machine learning algorithms offer promising accuracy rates, particularly CNNs and ensemble models, in diagnosing PCOS using various non-invasive parameters. Understanding phenotypic variations aids in tailoring treatments based on specific subgroups, potentially improving patient outcomes and management. Efforts toward identifying early markers or predictive factors can facilitate early intervention and lifestyle modifications, mitigating long-term health risks associated with PCOS.

##### Drawbacks and Recommendations:
Some studies may suffer from limited sample sizes or datasets, impacting the generalizability of findings. Larger and more diverse datasets are recommended for robust model development and validation. While certain machine learning models showcase high accuracy, the variability in dataset characteristics and preprocessing methods could influence their effectiveness across different populations or settings. The multifaceted nature of PCOS, influenced by both genetic and environmental factors, presents challenges in pinpointing a singular cause or standard diagnostic criteria.

The collective body of literature reviews signifies advancements in PCOS diagnosis, understanding clinical manifestations, and potential avenues for tailored treatments. The use of machine learning models, especially those employing CNNs and ensemble methods, showcases significant promise in accurate PCOS diagnosis based on non-invasive parameters. However, further research is necessary, emphasizing larger and more diverse datasets, refined models, and a multidisciplinary approach to fully comprehend the complexity of PCOS and enhance diagnostic and treatment strategies.


$~$

## Dataset:

The Polycystic ovary syndrome (PCOS) dataset, available on [Kaggle.com](https://www.kaggle.com/datasets/prasoonkottarathil/polycystic-ovary-syndrome-pcos), is comprised of two csv files labeled `PCOS_data_without_infertility` and `PCOS_infertility`. In total, these files encompass 48 variables and 541 data entries all collected from 10 different hospitals across Kerala, India. The dataset contains all physical and clinical parameters to determine PCOS and infertility related issues.

**Full description of the variables below:**

* Units used range from imperial to metric system of measurement

* For Yes | No questions
  * Yes = 1
  * No = 0

Variables                               Description
----------------------------            ---------------------------------------
"Sl..No"                                unique identification number assigned to each entry
"Patient.File.No."                      file number for each patient's record.       
"PCOS..Y.N."                            indicates the presence or absence of PCOS, with "Y" denoting "1 or Yes" and "N" for "0 or No."
"I...beta.HCG.mIU.mL."                  pregnancy hormone case I measured in milli-international units per liter (mIU/L)
"II....beta.HCG.mIU.mL."                pregnancy hormone case II measured in milli-international units per liter (mIU/L)
"AMH.ng.mL."                            detects ovarian reserve (egg count)           
"Age..yrs."                             age of patient in years            
"Weight..Kg."                           weight of patient in kg            
"Height.Cm."                            height of patient in cm            
"BMI"                                   body mass index                    
"Blood.Group"                           Blood Groups: A+ = 11, A- = 12, B+ = 13, B- = 14, O+ = 15, O- = 16, AB+ = 17, AB- = 18            
"Pulse.rate.bpm."                       beats per minute       
"RR..breaths.min."                      respiration rates per minute       
"Hb.g.dl."                              hemoglobin concentration measured in grams per deciliter (g/dL).               
"Cycle.R.I."                            cycle Regularity Index used to assess the regularity or irregularity of menstrual cycles in women: 4 indicates irregular menstrual cycle, 2 indicates a regular menstrual cycle           
"Cycle.length.days."                    length of menstrual cycle     
"Marraige.Status..Yrs."                 years married  
"Pregnant.Y.N."                         pregnant yes or no         
"No..of.aborptions"                     number of abortions      
"FSH.mIU.mL."                           follicle stimulating hormone measured in milli-international units per liter (mIU/L)            
"LH.mIU.mL."                            luteinizing hormone (increases during ovulation) measured in milli-international units per liter (mIU/L)             
"FSH.LH"                                ratio between Follicle-Stimulating Hormone (FSH) and Luteinizing Hormone (LH)                
"Hip.inch."                             measurement of hips in inches              
"Waist.inch."                           measurement of waist in inches            
"Waist.Hip.Ratio"                       ratio of measurement of waist and hip       
"TSH..mIU.L."                           thyroid stimulating hormone measured in milli-international units per liter (mIU/L)           
"AMH.ng.mL."                            Anti-Müllerian Hormone (AMH) measured in nanograms per milliliter (ng/mL); a marker used in reproductive medicine to assess ovarian reserve             
"PRL.ng.mL."                            Prolactin measured in nanograms per milliliter (ng/mL); a hormone produced by the pituitary gland           
"Vit.D3..ng.mL."                        Vitamin D3 measured in nanograms per milliliter (ng/mL); is essential for bone health, immune function, and various other bodily processes.         
"PRG.ng.mL."                            Progesterone measured in nanograms per milliliter (ng/mL); a hormone involved in the menstrual cycle, pregnancy, and maintaining the uterine lining for a developing embryo.             
"RBS.mg.dl."                            Random Blood Sugar measured in milligrams per deciliter (mg/dL); it represents the level of glucose (sugar) present in the blood at a random time, without fasting.            
"Weight.gain.Y.N."                      weight gain yes or no       
"hair.growth.Y.N."                      hair growth yes or no (hirsutism)       
"Skin.darkening..Y.N."                  darkening of skin yes or no  
"Hair.loss.Y.N."                        hair loss yes or no         
"Pimples.Y.N."                          pimples (acne) yes or no           
"Fast.food..Y.N."                       consumption of fast food yes or no    
"Reg.Exercise.Y.N."                     regularly exercise yes or no                 
"BP._Systolic..mmHg."                   systolic blood pressure measured in millimeters of mercury (mmHg).     
"BP._Diastolic..mmHg."                  diastolic blood pressure measured in millimeters of mercury (mmHg)  
"Follicle.No...L."                      number of follicles on left ovary       
"Follicle.No...R."                      number of follicles in right ovary       
"Avg..F.size..L...mm."                  average size of follicles in left ovary measured in millimeters (mm)  
"Avg..F.size..R...mm."                  average size of follicles in right ovary measured in millimeters (mm)  
"Endometrium..mm."                      size of the endometrial thickness in millimeters (mm)


$~$

## Methodology:


Data Collection and Preparation: Gather and preprocess a comprehensive dataset containing demographic details, clinical measurements (hormone levels, BMI, menstrual history), and lifestyle information (diet, exercise, stress levels) from a diverse population.

Feature Selection and Engineering: Identify the most relevant features through exploratory analysis and feature engineering techniques, ensuring that the model focuses on key predictors for PCOS.

Model Training and Evaluation: Train the machine learning model using appropriate algorithms (such as Logistic Regression, Random Forest, Support Vector Machines) on a subset of the dataset, validate its performance using cross-validation techniques, and evaluate its accuracy, precision, recall, and F1-score.

Prediction and Risk Assessment: Deploy the trained model to predict the likelihood or risk of PCOS in new, unseen data, providing valuable insights into individuals who might be predisposed to or already have the condition.

Interpretation and Recommendations: Interpret the model's findings, analyze the significance of various factors contributing to the prediction, and offer recommendations or interventions based on identified risk factors to potentially prevent or manage PCOS.

This project will investigate mainly publicly available datasets that will be used to create predictive models on markers in routine test results to make a diagnosis. Some variables that are included in these datasets are:

*	Age
*	Weight
*	BMI
*	Race/ethnicity
*	Family history of PCOS
*	Menstrual cycle irregularity
*	Hormone levels (e.g., testosterone, LH, FSH)
*	Insulin resistance
*	Physical activity level
*	Diet

I will be using R (statistical performing language) to perform exploratory data analysis to process and analyze the data to check for structural errors and be able to create graphs and perform tests with minimal errors. Once the data is ready to use, I will be splitting the data into a training and test set to be able to use a machine learning algorithm such as logistic regression to create a predictive model. The predictive model will be based on markers (variables mentioned above) used to identify individuals who are at high risk for PCOS and target interventions to manage the condition.

To answer my research question:
* The datasets publicly available and the NICHD Dash looking to obtain all have PCOS and non-PCOS patients including demographic information, medical history, and laboratory tests. Preprocess the data by removing missing values, outliers, and redundant variables. Perform feature selection to identify the most informative variables for prediction.
* Split the dataset into training, validation, and testing sets. The training set is used to train the machine learning algorithm, the validation set is used to tune hyperparameters and prevent overfitting, and the testing set is used to evaluate the performance of the final model.
* Select an appropriate machine learning algorithm for the task at hand, such as logistic regression, decision trees, random forests, support vector machines, or neural networks. Train the algorithm on the training set using various techniques, such as cross-validation and regularization, to optimize its performance.
* Evaluate the performance of the trained model on the validation set using various metrics, such as accuracy, precision, recall, F1 score, and area under the curve. Use feature importance analysis to identify the most influential variables for prediction.
* Tune the hyperparameters of the machine learning algorithm using grid search, random search, or Bayesian optimization to improve its performance on the validation set.
* Select the final model based on its performance on the validation set. Evaluate its performance on the testing set to assess its generalization ability.
* Interpret the results of the machine learning algorithm using various techniques, such as decision trees, feature importance analysis, and partial dependence plots. Visualize the results using graphs, charts, and heatmaps to facilitate understanding and communication.
* Deploy the trained model on new data and disseminate the findings through scientific publications, presentations, and online platforms.

Note: this methodology plan is not exhaustive and may vary depending on the specific research question, dataset, and machine learning algorithm used.

Note 2: data has already been collected, there is no need for me to gather participants, perform exams (such as bloodwork), use medical equipment to collect the data, perform surveys, have a location to perform a study, etc. I will be the sole person studying the data set and conducting the analysis.

$~$

## Assumptions:

While there’s limited information available in the medical field and even less data sets available to analyze, I have some concerns on being successful in predicting a PCOS diagnosis. Yet there are justifications exploring PCOS in depth:

* Identify diagnostic biomarkers that can distinguish PCOS patients from healthy individuals or those with other disorders. These biomarkers can aid in earlier diagnosis and better management of the disease.

* Predict the likelihood of disease progression and the risk of developing complications, such as diabetes and cardiovascular disease, in PCOS patients. This information can guide treatment decisions and improve patient outcomes.

* Develop personalized treatment plans for PCOS patients based on their individual characteristics and medical history. This approach can lead to more effective and targeted interventions.

* Integrate data from various sources, such as electronic health records, imaging studies, and genetic analyses, to provide a more comprehensive understanding of PCOS. This can help identify new pathways involved in the disease and potential targets for therapy.

* Aid in the design and analysis of clinical trials, leading to more efficient and informative studies. This can accelerate the development of new treatments for PCOS.

Early diagnosis and management of PCOS can lead to better health outcomes, improved quality of life, and reduced long-term health risks. Therefore, predicting PCOS diagnosis can have several societal benefits, including:

* Predicting PCOS diagnosis can help healthcare providers identify women at risk of developing PCOS and intervene early with appropriate treatment, such as lifestyle modifications and medication, to prevent or minimize the long-term health consequences of the disorder.

* Early diagnosis and treatment of PCOS can help manage symptoms such as irregular periods, infertility, acne, and excess hair growth, leading to improved physical and mental health outcomes for affected women.

* By predicting PCOS diagnosis and intervening early, healthcare providers can prevent or reduce the need for more expensive treatments or surgeries later in life, resulting in cost savings for individuals, healthcare systems, and society.

* Predicting PCOS diagnosis can increase awareness of the disorder among healthcare providers, patients, and the public, leading to more education, research, and advocacy efforts aimed at improving PCOS diagnosis, treatment, and management.

* Early intervention and management of PCOS can improve the quality of life for affected women, leading to increased productivity, better mental health, and greater overall well-being.

Overall, I'll be able to explore the insights into PCOS pathophysiology, diagnosis, and treatment. Their use in PCOS research can lead to more personalized and effective care for patients with this complex disorder.


$~$

## Experimentation and Results:

$~$

## Conclusion:


My initial assumption was that while there’s limited information available in the medical field and even less data sets available to analyze, I have some concerns on being successful in predicting a PCOS diagnosis. 

$~$

## References:

1.	Bulsara, J., Patel, P., Soni, A., &amp; Acharya, S. (2021, February 10). A review: Brief insight into polycystic ovarian syndrome. Endocrine and Metabolic Science. Retrieved February 23, 2023, from https://www.sciencedirect.com 

2.	World female population, 1960-2022. Knoema. (2022). Retrieved February 24, 2023, from https://knoema.com. 

3.	MarchofDimes. (2022). Population of women 15-44 years by age: United States, 2020. March of Dimes | PeriStats. Retrieved February 24, 2023, from https://www.marchofdimes.org

4.	Engmann, L., Jin, S., Sun, F., Legro, R. S., Polotsky, A. J., Hansen, K. R., Coutifaris, C., Diamond, M. P., Eisenberg, E., Zhang, H., Santoro, N., & Reproductive Medicine Network (2017). Racial and ethnic differences in the polycystic ovary syndrome metabolic phenotype. American journal of obstetrics and gynecology, 216(5), 493.e1–493.e13. https://doi.org

5.	Fehring, Richard J., "Menstrual Cycle Data" (2012). Randomized Comparison of Two Internet-Supported Methods of Natural Family Planning. 7.
https://epublications.marquette.edu

6.	Khan, M. J., Ullah, A., & Basit, S. (2019). Genetic Basis of Polycystic Ovary Syndrome (PCOS): Current Perspectives. The application of clinical genetics, 12, 249–260. https://doi.org/


$~$

## Appendices:

### Appendix A - Figures:

```{r,  out.width="80%", fig.align="center", warning=FALSE, message=FALSE, echo=FALSE}

```


$~$

### Appendix B - Tables:

```{r, out.width="80%", fig.align="center", warning=FALSE, message=FALSE, echo=FALSE}

```


$~$

### Appendix C - R Code:

```{r, eval=FALSE}

```

